\pchapter{Installation} \label{ch:Installation}
The \PPODESUITE~package depends on:
\begin{description}
  \item[\MATLAB] Tested on \MATLAB~version R2013b (8.2) 64-bit.
  \item[gfortran and GCC] Tested on version 4.8.1 64-bit.
\end{description}

\psection{Linux and other Unix variants}
\begin{enumerate}
 \item Meet the software requirements by installing \MATLAB~and the GCC package. \MATLAB~download and installation instructions can be found on the \MATHWORKS~website. The gfortran/GCC package can be obtained from the \nix~distribution repository through the distribution package manager. Using \UBUNTU~for example:
 \shellcmd{sudo apt-get install gfortran gcc}
\end{enumerate}

\psection{\MSDOS}
\textbf{Currently, running the \PPODESUITE~on \MSDOS~is not supported nor tested.}
\begin{enumerate}
 \item Running the \PPODESUITE~on \MSDOS~would require installing a combination of GCC/Cygwin and gnumex (\hyperlink{http://gnumex.sourceforge.net/}) or the Intel Visual Fortran Composer. To check which versions are supported by \MATLAB, check \hyperlink{http://www.mathworks.nl/support/compilers/R2013b/} substituting R2013b with the right version number.
\end{enumerate}

\psection{General}
\begin{enumerate}[resume]
 \item Download and extract the \PPODESUITE~package.
 \item Open matlab and navigate to the extracted \PPODESUITE~folder. Add the \PPODE~paths to the matlab path variable.
 \matlabcmd{PPODE\_addPaths}
 \item Now the libraries of the different solvers can be build. In order to do so, execute the following command;
 \matlabcmd{PPODE\_init}
 The options 'Debug' can be used to build the libraries with debugging symbols.
 \matlabcmd{PPODE\_init('Debug', 1)}
 For more information, type \matlabcmdinline{help PPODE\_init} in the \MATLAB~command window.
\end{enumerate}
